Refractory anaphylaxis: Data from the European Anaphylaxis Registry

### 2018-09-18 analysis/paper/paper.Rmd
corrections after final revision

# Abstract
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction. Comprehensive data on its definition, prevalence, and risk factors are missing.  
Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analysed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).

The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`). The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases it was applied intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options (e.g. vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases. The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).

Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved. The lack of second line medication (i.e. methylene blue or dopamine) use in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.

@@

Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering and augmenting factors [@Simons2015]. This variability introduces a wide range of possible reaction-symptom severities. Therefore multiple sub-types of anaphylaxis have been previously identified (i.e. food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).

@@

<!-- ## Data handling and statisitcs -->
The statistical analysis was performed in the R Statistical Package [@R]. Simple comparison of nominal variables was performed using Fischer's exact test, continous variables were analysed using Mann-Whithney U test. We defined statistical significance as α = 0.05. Data along with the analysis script can be accessed at github.com/wolass/RefractoryAnaOrg.

@@

## Refractory anaphylaxis is life-threathening
<!-- ## Symptoms -->
Milder anaphylaxis symptoms (i.e. pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases. Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
<!-- ## Death  -->
Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful. In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g. hypoxic brain injury), death occured in the next 3-8 days.

@@

Corticosteroids i.v. were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases. Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases but was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment. 
<!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->

@@

## Frequency and patients at risk
Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines [@Muraro2014]. Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly. An early use of drugs (e.g. methylene blue and dopamine) might increse the survival chance.

Overall, the severity of an anaphylaxis and its probable transition into a refractory episode might be described by four co-influencing mechanisms: 
1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolisation; 
2) cofactors - the presence of other factors which may increase the severity of a given reaction i.e. concomitant use of a betablocker etc.;
3) compensation - how well is the body able to compensate reaction symptoms e.g. hypotension; 
4) therapy - how fast and accurately the therapeutic procedures were introduced (Fig. \@ref(fig:theory)).
<!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->

## Elicitors
Drugs were the most commonly occurring elicitor, which is in concordance with the literature and our previous findings [@Francuzik2018]. Multiple medications with mast cell activation potential [@Hepner2003] given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007]. Recent study on IgE-independent anaphylaxis showed that peroperative drugs (i.e. rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MrgprX2 receptor [@Mcneil2015]. Therefore patients who undergo surgical procedures and have a history of anaphylaxis should remain under extraordinary caution.

@@

Only few refractory cases were elicited by yellow-jacket-stings. Therefore, the ratio of yellow-jacket to honey-bee venom elicited anaphylaxis was inverted in the refractory group. Reason for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
<!-- Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) . -->

@@

## Risk factors 
Although we have previously identified older age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered. Asthma and malignant diseases in the medical history were more frequently present in the refractory anaphylaxis group. The underlying reason is unclear. It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.

<!-- Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis? -->
```{r}
temp <- data3[data3$rANA=="yes",]
mean_age_malig <- data3$d_age %>% split(data3$q_410_malig_prev_v5 %>% factor) %>% lapply(mean)
```
Many cases of hypersenitivity due to cancer drugs were described in the literature, but malignancy in medical history has never been previously associated with severe anaphylaxis. 
It can be possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
<!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
0.001) and therefore less able to compensate the anaphylactic shock.

@@

Psychological stress in temporal proximity to the reaction might increase the severity of the reaction as we previously reported [@Worm2018]. It has been reported that stress induces the complement cascade activation [@Burns2008]. Substance P which is released during stress [@Arck2003] is a known vasodilator [@Bos92] and can activate mast cells directly [@Singh1999]. These mediators may contribute to an increased mast cell response and more severe anaphylaxis.

@@

Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases. Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the \beta-adrenergic receptor blockade [@Zaloga1986].

@@

Definition  of refractory anaphylaxis is not universal. Also in this study, the confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although it in fact could have been refractory.

@@

```{r demoTab}
knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other variables as fractions [%]. DM - diabetes mellitus, p value is derived from a Mann-Whitney U test or a Chi^2^ test for interval and categorical variables respectively.",
             digits = 2,
             align = rep("r",10),
             nsmall=2)
```

@@

```{r elicitorTab}
knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test.",
             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]")
            #digits = 1,
             # align = rep("r",10),
             # trim=T,
             # scientific=F,
             # nsmall=3,
             #drop0trailing = F
             )
```

@@

```{r elicitExact}
knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
```

p-value derived from the Fisher exact test.

@@

```{r symptTab}
knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"))
```

@@

```{r therapyTab}
knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. 2nd line therapy was performed after the initial rescue individual called for a professional medical emergeny team. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
```

@@

```{r cofactorsTab}
knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"))
```

>>>>

# Abstract
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction. Comprehensive data on its definition, prevalence, and risk factors are missing. Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analysed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).

The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`). The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases it was applied intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options (e.g. vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases. The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).

Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved. The neglegence of second line medication (i.e. methylene blue or dopamine) in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.

@@

Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering and augmenting factors [@Simons2015]. This variability introduces a wide range of possible reaction-symptom severities. Therefore, multiple sub-types of anaphylaxis have been previously identified (i.e. food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).

@@

<!-- ## Data handling and statisitcs -->
The statistical analysis was performed in the R Statistical Package [@R]. Simple comparison of categorical variables was performed using Fisher's exact test, continous variables were analysed using Mann-Whithney U test. We defined statistical significance as α = 0.05. Data along with the analysis script can be accessed at www.github.com/wolass/RefractoryAnaOrg.

@@

## Refractory anaphylaxis is life-threathening
<!-- ## Symptoms -->
Milder anaphylaxis symptoms (i.e. pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases. Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
<!-- ## Death  -->
Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful. In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g. hypoxic brain injury), death occured in the next 3-8 days.

@@

Corticosteroids i.v. were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases. Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases and was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment. 
<!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->

@@

## Frequency and patients at risk
Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines [@Muraro2014]. Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly. An early use of drugs (e.g. methylene blue and dopamine) might increse their survival chance.

Overall, the severity of an anaphylaxis and its probable transition into a refractory episode might be described by four co-influencing mechanisms: 
1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolism; 
2) cofactors - the presence of other factors which may increase the severity of a given reaction i.e. concomitant use of a betablocker etc.;
3) compensation - how well is the body able to compensate reaction symptoms e.g. hypotension; 
4) therapy - how fast and accurately the therapeutic procedures were introduced (Fig. \@ref(fig:theory)).
<!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->

## Elicitors
Drugs were the most commonly occurring elicitor, which is in concordance with the literature and our previous findings [@Francuzik2018]. Multiple medications with mast cell activation potential [@Hepner2003] given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007]. Recent study on IgE-independent anaphylaxis showed that peroperative drugs (i.e. rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MRGPRX2 receptor [@Mcneil2015]. Therefore patients who undergo surgical procedures and have a history of anaphylaxis should remain under extraordinary caution.

@@

Only few refractory cases were elicited by yellow-jacket-stings. Therefore, the ratio of yellow-jacket to honey-bee venom elicited anaphylaxis was inverted in the refractory group. Reasons for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
<!-- Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) . -->

@@

## Risk factors 
Although we have previously identified older age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered. Asthma and malignant diseases in the medical history were more frequently present in the refractory anaphylaxis group. The underlying reason is unclear.

<!-- Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis? -->
```{r}
temp <- data3[data3$rANA=="yes",]
mean_age_malig <- data3$d_age %>% split(data3$q_410_malig_prev_v5 %>% factor) %>% lapply(mean)
```
It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients. Many cases of hypersenitivity due to cancer drugs were described in the literature, but malignancy in medical history has never been previously associated with severe anaphylaxis. It is possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
<!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
0.001) and therefore less able to compensate the anaphylactic shock.

@@

Psychological stress in temporal proximity to the reaction might increase the severity of the reaction as we previously reported [@Worm2018]. Stress induces the complement cascade activation [@Burns2008]. Substance P, which is released during stress [@Arck2003], is a known vasodilator [@Bos92] and can activate mast cells directly [@Singh1999]. These mediators may contribute to an increased mast cell response and more severe anaphylaxis.

@@

Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases. Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the β-adrenergic receptor blockade [@Zaloga1986].

@@

Definition  of refractory anaphylaxis is not universal. The confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it in fact could have been refractory.

@@

```{r demoTab}
knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other variables as fractions [%]. DM - diabetes mellitus, p value is derived from a Mann-Whitney U test or a Chi^2^ test for interval and categorical variables respectively. Refractory group was compared to severe anaphylaxis cases without differentiating into male and female subgroups.",
             digits = 2,
             align = rep("r",10),
             nsmall=2)
```

@@

```{r elicitorTab}
knitr::kable(elicitorTab[,c(1,5,6,7,8,2,3,4)], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test.",
             col.names = c("n","Age","Male sex [%]","Perioperative [%]","Food allergy [%]","refractory ANA [%]","severe ANA [%]","p value")
            #digits = 1,
             # align = rep("r",10),
             # trim=T,
             # scientific=F,
             # nsmall=3,
             #drop0trailing = F
             )
```

@@

```{r elicitExact}

knitr::kable(cbind(c("Antibiotics","X-ray medium","Muscle relaxant","Legumes","Bee venom","Yellow-jacket venom"),elicitExact[,2:4]), caption = "Summary of the elicitors broken down into specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
```

@@

```{r symptTab}
knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"))
```

@@

```{r therapyTab}
knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. '2nd line' refers to the therapy performed by the professional medical emergency team after the initial rescue procedures. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
```

@@

```{r cofactorsTab}
knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases. ANA - anaphylaxis, p-value derived from the Fisher's exact test. ",
             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"))
```
